---
title: "Autoimmune liver diseases UKB"
author: "Jesper Sloth Kellemann and Anne-Sofie H. Jensen"
date: "2026-08-05"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r echo=T, results='hide', warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(ukbtools)
library(data.table)
library(dplyr)
df <- fread(here("data-raw/ukb_all_datasets.csv"), data.table = FALSE)
getwd()
nrow(df) # 502488
```

### Remove individuals who have withdrawn
```{r echo=T, results='hide', message=FALSE, warning=FALSE}
withdraw <- read_csv(here("data-raw/withdrawal_2024-12-17.csv"), col_names = "eid") 
df <- anti_join(df, withdraw, by = "eid")
nrow(df) # 502133
```

### Retrieving ICD diagnoses
```{r}
ukb_icd_code_meaning(icd.code = c("K732", "K754", "K738", "K743", "K70", "K760","K703","B180","B181","B182","K746","I850","I859","I982","I983","K766","K767","K729",
                                  "I10","I110","I119","I120","I129","I130","I131","I132","I139","I150","I151","I152","I158","I159","E10","E11","E14"), icd.version = 10)
```

### Import ICD10 dates dataframe and merge
df_icd contain data from our project "Olink All proteins" on DNA nexus from 13.02.2024.
df contain data from UKB participants without those who have withdrawn up until 17.12.2024 (see name of withdrawal file); therefore fewer eids.
```{r echo=TRUE, results = 'hide'}
df_icd <- fread(here("data-raw/ICD10_dates_Fie.csv"), data.table = FALSE)
dim(df_icd) # 502202 
df_icd[1:3,]
dim(df) # 502133

# Check if all "eid" rows in df are present in the "eid" column of df_icd
all(df$eid %in% df_icd$eid)

# Merge the dataframes by 'eid', keeping only those eids present in both dataframes
df <- merge(df_icd, df, by = "eid", all = FALSE)
dim(df) # 502131
```

### Import dataset with extra liver test variables
lt contain data from our project "Olink All proteins" on DNA nexus from 13.02.2024 (like df_icd and now df)
ALT, AST, ALP, GGT are U/L in UKB. Bilirubin is umol/L
```{r echo=TRUE, results = 'hide'}
lt <- fread(here("data-raw/livertests.csv"), data.table = FALSE)
dim(lt)
lt[1:3,]
df$alanine_aminotransferase_f30620_0_0
# Rename variables
lt$ALP_base <- lt$p30610_i0
lt$bili_base <- lt$p30840_i0
lt$AST_base <- lt$p30650_i0
lt$GGT_base <- lt$p30730_i0
dim(lt)
lt[1:3,]

# Subset the lt dataframe
lt_subset <- lt %>% select(eid, ALP_base, bili_base, AST_base, GGT_base)
dim(lt)
dim(lt_subset)
lt_subset[1:3,]
```

### Merge lt dataset and df dataset by eid
```{r echo=T, results='hide'}
dim(lt_subset) # 502202
dim(df) # 502131
all(df$eid %in% lt_subset$eid) # same eids

# Merge the dataframes by 'eid'
dim(df)
dim(lt_subset)
df <- merge(lt_subset, df, by = "eid", all = FALSE)
dim(df) # 502131
```

### Import enrollment date
enroll contain data from our project "Olink All proteins" on DNA nexus from 13.02.2024 (like df_merge)
```{r echo=TRUE, results = 'hide'}
enroll <- fread(here("data-raw/enrollemnt_date.csv"), data.table = FALSE)
dim(enroll) # 502202
enroll[1:3,]
enroll$enroll_date <- enroll$p53_i0
enroll <- enroll[,c(1,3)]
enroll[1:3,]
```

### Merge enroll dataset and df dataset by eid
```{r echo=TRUE, results = 'hide'}
dim(df)
dim(enroll)
df <- merge(df, enroll, by = "eid", all = FALSE)
dim(df)
df[1:3,947:949]
```

### Create diagnoses TRUE/NA
Autoimmune liver diseases and cirrhosis
```{r echo=T, results='hide'}

# New AIH column and make AIH TRUE if not NA in date column K732, K738, K754
df$AIH <- NA
aih_non_na_dates <- df %>%
  filter(!is.na(date_K732) | !is.na(date_K738) | !is.na(date_K754)) %>%
  pull(eid)

# Update the AIH column
df <- df %>%
  mutate(AIH = ifelse(eid %in% aih_non_na_dates, TRUE, AIH))

# Check the updated AIH column
table(df$AIH, useNA = "ifany") # 444 cases

# New PBC column and make PBC TRUE if not NA in date column K743
df$PBC <- NA
pbc_non_na_dates <- df %>%
  filter(!is.na(date_K743)) %>%
  pull(eid)

# Update the PBC column
df <- df %>%
  mutate(PBC = ifelse(eid %in% pbc_non_na_dates, TRUE, PBC))

# Check the updated PBC column
table(df$PBC, useNA = "ifany") # 451

# Make individuals having both AIH and PBC (overlap syndrome) into a new variable (AIH.PBC)
df %>%
      count(AIH,PBC) # 45 with overlap

df$AIH.PBC[df$AIH==T & df$PBC ==T] <- TRUE
df$AIH[df$AIH.PBC == T] <- NA
df$PBC[df$AIH.PBC == T] <- NA

df %>%
      count(AIH,PBC,AIH.PBC) # 399, 406, 45 respectively

# Add chronic hep B
df$HBV <- NA
hbv_non_na_dates <- df %>%
  filter(!is.na(date_B180) | !is.na(date_B181)) %>%
  pull(eid)

# Update the HBV column
df <- df %>%
  mutate(HBV = ifelse(eid %in% hbv_non_na_dates, TRUE, HBV))

# Check the updated HBV column
table(df$HBV, useNA = "ifany") # 365 cases

# Add chronic hep C
df$HCV <- NA
hcv_non_na_dates <- df %>%
  filter(!is.na(date_B182)) %>%
  pull(eid)

# Update the HCV column
df <- df %>%
  mutate(HCV = ifelse(eid %in% hcv_non_na_dates, TRUE, HCV))

# Check the updated HCV column
table(df$HCV, useNA = "ifany") # 521 cases

# Make those with AIH that have concurrent HBV/HCV as AIH-NA (i.e. not having AIH)

df %>%
      count(AIH,HBV,HCV) # 26 individuals

df$AIH[df$HBV == T & is.na(df$HCV)] <- NA
df$AIH[is.na(df$HBV) & df$HCV == T] <- NA
df$AIH[df$HBV == T & df$HCV == T] <- NA

table(df$AIH) # 373 individuals

# Same with PBC and concurrent HBV/HCV

df %>%
      count(PBC,HBV,HCV)  # 3 individuals

df$PBC[df$HBV == T & is.na(df$HCV)] <- NA
df$PBC[is.na(df$HBV) & df$HCV == T] <- NA

table(df$PBC) # 403

# None with AIH.PBC and concurrent HBV/HCV

df %>%
      count(AIH.PBC,HBV) 

df %>%
      count(AIH.PBC,HCV) 

# Cirrhosis
df$cir <- NA
cir_non_na_dates <- df %>%
  filter(!is.na(date_K746) | !is.na(date_I850) | !is.na(date_I859) | !is.na(date_I982) | !is.na(date_I983) | !is.na(date_K729) | !is.na(date_K703) |
             !is.na(date_K766) | !is.na(date_K767)) %>%
  pull(eid)

# Update the cir column
df <- df %>%
  mutate(cir = ifelse(eid %in% cir_non_na_dates, TRUE, cir))
```

Calculate sum of average weekly intake of alcohol
```{r echo=TRUE, results='hide'}
# Set prefer not to answer and do not know (-1 and -3) to NA
df$average_weekly_beer_plus_cider_intake_f1588_0_0[df$average_weekly_beer_plus_cider_intake_f1588_0_0<0] <- NA
df$average_weekly_champagne_plus_white_wine_intake_f1578_0_0[df$average_weekly_champagne_plus_white_wine_intake_f1578_0_0<0] <- NA
df$average_weekly_fortified_wine_intake_f1608_0_0[df$average_weekly_fortified_wine_intake_f1608_0_0<0] <- NA
df$average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0[df$average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0<0] <- NA
df$average_weekly_red_wine_intake_f1568_0_0[df$average_weekly_red_wine_intake_f1568_0_0<0] <- NA
df$average_weekly_spirits_intake_f1598_0_0[df$average_weekly_spirits_intake_f1598_0_0<0] <- NA

# units of alcohol in beverages: https://www.nhs.uk/live-well/alcohol-advice/calculating-alcohol-units/

# weekly units of alcohol - redwine (f1568), 1 glass = 2.1 units
df$average_weekly_red_wine_intake_f1568_0_0 <- df$average_weekly_red_wine_intake_f1568_0_0 * 2.1
# weekly units of alcohol - champagne etc (f1578), 1 glass = 2.1 units
df$average_weekly_champagne_plus_white_wine_intake_f1578_0_0 <- df$average_weekly_champagne_plus_white_wine_intake_f1578_0_0 * 2.1
# weekly units of alcohol - beer and cider (f1588), 1 pint = 2 units
df$average_weekly_beer_plus_cider_intake_f1588_0_0 <- df$average_weekly_beer_plus_cider_intake_f1588_0_0 * 2
# weekly units of alcohol - spirits (f1598), 1 measure = 1 unit (no calculation needed)
# weekly units of alcohol - fortified wine (f1608), 1 measure = 1 unit (no calculation needed)
# weekly units of other alcoholic drinks incl alcopops, 1 glass = 1.5 units
df$average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0 <- df$average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0 * 1.5

# Calculate the sum of weekly units alcoholic intake
df <- df %>%
    rowwise() %>%
    mutate(weekly_alcohol = sum(average_weekly_beer_plus_cider_intake_f1588_0_0,
    average_weekly_champagne_plus_white_wine_intake_f1578_0_0,
    average_weekly_fortified_wine_intake_f1608_0_0,
    average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0,
    average_weekly_red_wine_intake_f1568_0_0,
    average_weekly_spirits_intake_f1598_0_0,
    na.rm=T))
```

Defining MASLD
Based on Rinella et al., JHEP, 2023 
```{r echo =TRUE, results='hide'}

# Identify source and timing of MASLD classification
# New masld_icd column and make masld_icd TRUE if not NA in date column K760
df$masld_icd <- NA
masld_icd_non_na_dates <- df %>%
  filter(!is.na(date_K760)) %>%
  pull(eid)

# Update the masld_icd column
df <- df %>%
  mutate(masld_icd = ifelse(eid %in% masld_icd_non_na_dates, TRUE, masld_icd))

# Create masld_icd_prev and masld_icd_inc variables based on timing of K760 diagnosis
df$masld_icd_prev <- with(df,
                          ifelse(!is.na(date_K760) & date_K760 <= enroll_date,
                                 TRUE, FALSE))

df$masld_icd_inc <- with(df,
                         ifelse(!is.na(date_K760) & date_K760 > enroll_date,
                                TRUE, FALSE))

# New masld_pdff column and make masld_pdff TRUE if PDFF and alcohol criteria are met
df$masld_pdff <- NA

df$masld_pdff[df$proton_density_fat_fraction_pdff_f40061_2_0 > 5.5 &
                 df$sex_f31_0_0 == "Female" &
                 df$weekly_alcohol < 17.5] <- TRUE

df$masld_pdff[df$proton_density_fat_fraction_pdff_f40061_2_0 > 5.5 &
                 df$sex_f31_0_0 == "Male" &
                 df$weekly_alcohol < 26.25] <- TRUE

# Check source and timing of MASLD classification
table(df$masld_icd, useNA = "ifany")
table(df$masld_pdff, useNA = "ifany")
table(df$masld_icd_prev, useNA = "ifany")
table(df$masld_icd_inc, useNA = "ifany")
table(df$masld_icd, df$masld_pdff, useNA = "ifany")


# New MASLD column and make MASLD TRUE if not NA in date column 
df$MASLD <- NA
masld_non_na_dates <- df %>%
  filter(!is.na(date_K760)) %>%
  pull(eid)

# Update the MASLD column
df <- df %>%
  mutate(MASLD = ifelse(eid %in% masld_non_na_dates, TRUE, MASLD))

# Number of PDFF measures in dataset with 40534 variable
table(is.na(df$proton_density_fat_fraction_pdff_f40061_2_0))

# Add pdff > 5.5 for women and men based on alcohol intake (MASLD when drinking less than 140 g alcohol per week for women and less than 210 g alcohol per week for men. 1 unit = 8 g alcohol => NAFLD when less than < 17.5 units alcohol/week for women and 26.25 units alcohol/week for men
df$MASLD[df$proton_density_fat_fraction_pdff_f40061_2_0 > 5.5 &
                 df$sex_f31_0_0 == "Female" &
                 df$weekly_alcohol < 17.5] <- T
df$MASLD[df$proton_density_fat_fraction_pdff_f40061_2_0 > 5.5 &
                 df$sex_f31_0_0 == "Male" &
                 df$weekly_alcohol<26.25] <- T

table(df$MASLD, useNA = "ifany") # 14183
```

Check if MASLD criteria are met and if not, make those individuals cryptogenic SLD
```{r echo =TRUE, results='hide', warning=FALSE, message=FALSE}
# Create a new column that categorizes BMI as "Below 25" and "25 or Above"
df$bmi_new <- ifelse(df$body_mass_index_bmi_f21001_0_0 < 25, "Below 25", "25 or Above")
table(df$MASLD, df$bmi_new, useNA = "ifany") # 1724 with a bmi below 25

# Create new crypt_sld variable with "lean MASLD" (trying to find those MASLD individuals, if any, who do not fulfill MASLD criteria)
df$crypt_sld <- ifelse(df$MASLD == TRUE & df$bmi_new == "Below 25",TRUE, NA)
table(df$crypt_sld, useNA = "ifany") # 1724

# Create a new column that categorizes HDL as "low" if HDL is 1 or below for Male and 1.3 or below for Female, else "high" (HDL is mmol/L in UKB)
df$hdl_new <- ifelse((df$hdl_cholesterol_f30760_0_0 <= 1 & df$hdl_cholesterol_f30760_1_0 <= 1 & df$sex_f31_0_0 == "Male") | 
                     (df$hdl_cholesterol_f30760_0_0 <= 1.3 & df$hdl_cholesterol_f30760_1_0 <= 1.3 & df$sex_f31_0_0 == "Female"), 
                     "low", "high")
table(df$crypt_sld, df$hdl_new, useNA = "ifany") 

# Make those crypt_sld TRUE having HDL low NA (meaning they have truly MASLD)
df$crypt_sld[df$hdl_new == "low"] <- NA
table(df$crypt_sld, useNA = "ifany") # 1724 - 16 = 1708

# Create a new column that categorizes triglyceride as low if below 1.7, else high (TG is mmol/L in UKB)
df$tg_new <- ifelse(df$triglycerides_f30870_0_0 < 1.7 & df$triglycerides_f30870_1_0 < 1.7, "low", "high")
table(df$crypt_sld, df$tg_new, useNA = "ifany")

# Make those crypt_sld TRUE having TG high NA (meaning they have truly MASLD)
df$crypt_sld[df$tg_new == "high"] <- NA
table(df$crypt_sld, useNA = "ifany") # 1708 - 673 = 1035

# Define the list of lipid-lowering medication codes
lipid_drug_codes <- c(
  # Statins
  1140861958, # simvastatin
  1140881748, # zocor (simvastatin)
  1141200040, # zocor heart-pro (simvastatin)
  1140888594, # fluvastatin
  1140864592, # lescol (fluvastatin)
  1140888648, # pravastatin
  1140910632, # eptastatin
  1140910654, # velastatin
  1141146234, # atorvastatin
  1141146138, # lipitor (atorvastatin)
  1141192410, # rosuvastatin
  1141192414, # crestor
  # Ezetimibe
  1141192736, # ezetimibe
  1141192740, # ezetimib
  # Niacin
  1140910670  # niacin
)

# Function to detect drugs in medication columns
find_lipid_drug_function <- function(df, codes) {
  # Select the relevant columns
  medication_cols <- df %>%
    select(starts_with("treatmentmedication_code_f20003"))
  
  # Check for matches and convert to a logical matrix
  match_matrix <- sapply(medication_cols, function(col) col %in% codes)
  
  # Create a logical vector for the oral lipid drug
  df$oral_lipid_drug <- rowSums(match_matrix, na.rm = TRUE) > 0
  return(df)
}

# Apply the function to the dataframe
df <- find_lipid_drug_function(df, lipid_drug_codes)

# View some rows where oral_lipid_drug is TRUE (checkup)
df %>%
  filter(oral_lipid_drug == TRUE) %>%
  select(eid, oral_lipid_drug, starts_with("treatmentmedication_code_f20003")) %>%
  head(10)  # Show first 10 such cases

table(df$crypt_sld, df$oral_lipid_drug, useNA = "ifany")

# Make those crypt_sld taking lipid lowering medication NA 
df$crypt_sld[df$oral_lipid_drug == TRUE] <- NA
table(df$crypt_sld, useNA = "ifany") # 1035 - 176 = 859

# New hypertension column and make HT TRUE if not NA in date column 
df$HT <- NA
ht_non_na_dates <- df %>%
  filter(!is.na(date_I10) | !is.na(date_I110) | !is.na(date_I119) | !is.na(date_I120) | !is.na(date_I129) | !is.na(date_I130) | !is.na(date_I131)
         | !is.na(date_I132) | !is.na(date_I139) | !is.na(date_I150) | !is.na(date_I151) | !is.na(date_I152) | !is.na(date_I158) | !is.na(date_I159)) %>%
  pull(eid)

# Update the HT column
df <- df %>%
  mutate(HT = ifelse(eid %in% ht_non_na_dates, TRUE, HT))

table(df$crypt_sld,df$HT, useNA = "ifany") 

# Make those crypt_sld having hypertension NA 
df$crypt_sld[df$HT == TRUE] <- NA
table(df$crypt_sld, useNA = "ifany") # 859 - 210 = 649

# Create a new column that categorizes HbA1c as low if below 39, else high (hba1c is mmol/mol in UKB)
df$hba1c_new <- ifelse(df$glycated_haemoglobin_hba1c_f30750_0_0 < 39 | df$glycated_haemoglobin_hba1c_f30750_1_0 < 39, "low", "high")
table(df$crypt_sld, df$hba1c_new, useNA = "ifany") # all low -> still TRUE crypt_sld

# Now diabetes. First correct classification of T1D and T2D (e.g. several individuals with a diagnosis of both T1D and T2D). 
# New T2D column and make T2D TRUE if not NA in date column 
df$T2D <- NA
t2d_non_na_dates <- df %>%
  filter(!is.na(date_E11) | !is.na(date_E14)) %>%
  pull(eid)

# Update the T2D column
df <- df %>%
  mutate(T2D = ifelse(eid %in% t2d_non_na_dates, TRUE, T2D))

# New T1D column and make T1D TRUE if not NA in date column 
df$T1D <- NA
t1d_non_na_dates <- df %>%
  filter(!is.na(date_E10)) %>%
  pull(eid)

# Update the T1D column
df <- df %>%
  mutate(T1D = ifelse(eid %in% t1d_non_na_dates, TRUE, T1D))

# Remove diabetes diagnosis if gestational diabetes only
# df$gdmonly <- NA
df$gdmonly[df$gestational_diabetes_only_f4041_0_0 == 1 | df$gestational_diabetes_only_f4041_1_0 == 1 | df$gestational_diabetes_only_f4041_2_0 == 1 | df$gestational_diabetes_only_f4041_3_0 == 1] <- T
df$T1D[df$gdmonly == T & df$T2D == T] <- NA
df$T2D[df$gdmonly == T & is.na(df$T1D)] <- NA
df$T1D[df$gdmonly == T & is.na(df$T2D)] <- NA

# Filter for oral antihyperglycemic drugs and insulin
# Define the list of oral T2D medication codes (excl metformin which is not restricted to only T2D use)
t2d_drug_codes <- c(
  # Sulfonylureas
  1140874718, 1140874744, 1140874746, 1141152590, 1141156984,
  1140874646, 1141157284, 1140874652, 1140874674, 1140874728,
  
  # Others
  1140868902, 1140868908, 1140857508, 1140882964,
  
  # Metglitinides
  1141173882, 1141173786, 1141168660,
  
  # Glitazones
  1141171646, 1141171652, 1141153254, 1141177600, 1141177606)

# Function to detect oral T2D drugs in medication columns
find_t2d_drug_function <- function(df, codes) {
  # Select the relevant columns
  medication_cols <- df %>%
    select(starts_with("treatmentmedication_code_f20003"))
  
  # Check for matches and convert to a logical matrix
  match_matrix <- sapply(medication_cols, function(col) col %in% codes)
  
  # Create a logical vector for the oral T2D drugs
  df$oral_t2d_drug <- rowSums(match_matrix, na.rm = TRUE) > 0
  return(df)
}

# Apply the function to the dataframe
df <- find_t2d_drug_function(df, t2d_drug_codes)

# Define the list of Metformin medication codes
metformin_codes <- c(
  1140884600,  # metformin
  1140874686,  # glucophage 500 mg tablet
  1141189090   # rosiglitazone 1mg / metformin 500mg tablet
)

# Function to detect Metformin drugs in medication columns
find_metformin_function <- function(df, codes) {
  # Select the relevant columns
  medication_cols <- df %>%
    select(starts_with("treatmentmedication_code_f20003"))
  
  # Check for matches and convert to a logical matrix
  match_matrix <- sapply(medication_cols, function(col) col %in% codes)
  
  # Create a logical vector for Metformin use
  df$metformin_use <- rowSums(match_matrix, na.rm = TRUE) > 0
  return(df)
}

# Apply the function to the dataframe
df <- find_metformin_function(df, metformin_codes)

# Define the list of Insulin medication codes
insulin_codes <- c(
  1140883066  # insulin product
)

# Function to detect Insulin use in medication columns
find_insulin_function <- function(df, codes) {
  # Select the relevant columns
  medication_cols <- df %>%
    select(starts_with("treatmentmedication_code_f20003"))
  
  # Check for matches and convert to a logical matrix
  match_matrix <- sapply(medication_cols, function(col) col %in% codes)
  
  # Create a logical vector for Insulin use
  df$insulin_use <- rowSums(match_matrix, na.rm = TRUE) > 0
  return(df)
}

# Apply the function to the dataframe
df <- find_insulin_function(df, insulin_codes)

# Make those individuals wihtout diabetes diagnosis as T2D if hba1c above 48:
df$T2D[(df$glycated_haemoglobin_hba1c_f30750_0_0 > 48 | df$glycated_haemoglobin_hba1c_f30750_1_0 > 48) & (is.na(df$T2D) & is.na(df$T1D))] <-  T

# Make those individuals without diabetes diagnosis as T2D if oral T2D medication (excluding metformin):
df$T2D[df$oral_t2d_drug == T & (is.na(df$T2D) & is.na(df$T1D))] <- T

# Make those T1D individuals on oral T2D drug medication T2D:
df$T1D[(df$T2D == T & df$T1D == T) & (df$oral_t2d_drug == T | df$metformin_use == T)] <- NA
df$T2D[df$T1D == T & (df$oral_t2d_drug == T | df$metformin_use == T)] <- T
df$T1D[df$oral_t2d_drug == T | df$metformin_use == T] <- NA

# Make those diagnosed <20 y age as T1D
df$T2D[(df$age_diabetes_diagnosed_f2976_0_0 < 20 | df$age_diabetes_diagnosed_f2976_1_0 < 20 | df$age_diabetes_diagnosed_f2976_2_0 < 20 | df$age_diabetes_diagnosed_f2976_3_0 < 20 ) & df$T1D == T & df$T2D == T] <- NA

# Make those with double diagnosis as T1D if insulin treatment within 1 year of diagnosis and actual insulin treatment:
df$ins1y[df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_0_0 == 1 | df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_1_0 == 1 | df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_2_0 == 1 | df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_3_0 == 1] <- T 
df$ins1y[is.na(df$ins1y)] <- NA
df$T2D[df$T1D == T & df$T2D == T & (df$ins1y & df$insulin_treatment)] <- NA

# Making those with double diagnosis T1D and T2D as T2D
df$T1D[df$T1D == T & df$T2D == T] <- NA

df %>%
      count(T1D, T2D) # 790, 51608, respectively

table(df$crypt_sld, df$T2D, useNA = "ifany") 

# Make those crypt_sld having T2D NA 
df$crypt_sld[df$T2D == TRUE] <- NA
table(df$crypt_sld, useNA = "ifany") # 649 - 27 = 622 

table(df$MASLD, df$crypt_sld, useNA = "ifany") 

# Make those MASLD TRUE individuals who are crypt_sld TRUE NA
df$MASLD[df$crypt_sld == TRUE] <- NA
table(df$MASLD, useNA = "ifany") # 14183 - 622 = 13561
```

Defining ALD
```{r echo=TRUE, results='hide'}
# Add ALD K70
df$ALD <- NA
ald_non_na_dates <- df %>%
  filter(!is.na(date_K70)) %>%
  pull(eid)

# Update the ALD column
df <- df %>%
  mutate(ALD = ifelse(eid %in% ald_non_na_dates, TRUE, ALD))

# Add pdff > 5.5 for women and men based on alcohol intake (more than 350 g alcohol/week for women and 420 g alcohol/week for men. Between MASLD and ALD is MetALD (140-350 g alcohol for women and 210-420 g alcohol for men) which is not defined in this dataset
df$ALD[df$proton_density_fat_fraction_pdff_f40061_2_0 > 5.5 &
                 df$sex_f31_0_0 == "Female" &
                 df$weekly_alcohol >= 43.8] <- T
df$ALD[df$proton_density_fat_fraction_pdff_f40061_2_0 > 5.5 &
                 df$sex_f31_0_0=="Male" &
                 df$weekly_alcohol >= 52.5] <- T # 8 g alcohol in 1 unit
```

Healthy defined as bmi < 25, TG low, HDL high, HbA1c < 39, normal liver blood tests, no ICD diagnosis incl not having MASLD (as MASLD in this dataset is also identified on scan and not only ICD10), weakly alcohol consumption < 17.5 units alcohol/week for women and 26.25 units alcohol/week for men. In order to make healthy group as comparable to discovery heatlhy group as possible. 
```{r echo=TRUE, results = 'hide'}
# Create the alt_base_uln variable
df$alt_base_uln <- with(df, ifelse(
  (sex_f31_0_0 == "Female" & alanine_aminotransferase_f30620_0_0 < 45) |  
  (sex_f31_0_0 == "Male" & alanine_aminotransferase_f30620_0_0 < 70),    
  "normal",                                                              
  "elevated"                                                             
))

# Create the ast_uln variable
df$ast_uln <- with(df, ifelse(
  (sex_f31_0_0 == "Female" & AST_base < 35) |  
  (sex_f31_0_0 == "Male" & AST_base < 45),    
  "normal",                                   
  "elevated"                                  
))

# Create the alp_uln variable
df$alp_uln <- with(df, ifelse(
  ALP_base < 105,      
  "normal",           
  "elevated"          
))

# Create the ggt_uln variable
df$ggt_uln <- with(df, ifelse(
  (sex_f31_0_0 == "Female" & GGT_base < 75) |  
  (sex_f31_0_0 == "Male" & GGT_base < 115),    
  "normal",                                    
  "elevated"                                   
))

# Create the bili_uln variable
df$bili_uln <- with(df, ifelse(
  bili_base < 25,      
  "normal",          
  "elevated"          
))

# Create Healthy variable
df$Healthy <- with(df, ifelse(
  bmi_new == "Below 25" &                           
  tg_new == "low" &                                 
  hdl_new == "high" &                               
  hba1c_new == "low" &                              
  no_icd10 == TRUE &                                
  is.na(MASLD) &                                    
  ((sex_f31_0_0 == "Female" & weekly_alcohol < 17.5) |  
   (sex_f31_0_0 == "Male" & weekly_alcohol < 26.25)) &   
  alt_base_uln == "normal" &                        
  ast_uln == "normal" &                             
  ggt_uln == "normal" &                             
  alp_uln == "normal" &                             
  bili_uln == "normal",                             
  TRUE, NA
))

table(df$Healthy, useNA = "ifany")
dim(df)
df[1:3,974:978]
```

### Tidy up dataframe
```{r echo=TRUE, results='hide'}
# Make NA -> FALSE
df_final <- df %>%
    replace_na(list(
        AIH = FALSE,
        PBC = FALSE,
        AIH.PBC = FALSE,
        MASLD = FALSE,
        ALD = FALSE,
        crypt_sld = FALSE,
        cir = FALSE,
        Healthy = FALSE))

# Clean up dataframe
df_final <- df_final %>%
    select(eid = eid,
           Age = age_when_attended_assessment_centre_f21003_0_0,
           Sex = sex_f31_0_0,
           hba1c = glycated_haemoglobin_hba1c_f30750_0_0,
           BMI = body_mass_index_bmi_f21001_0_0,
           ethnicity = ethnic_background_f21000_0_0,
           ALT = alanine_aminotransferase_f30620_0_0,
           ALT1 = alanine_aminotransferase_f30620_1_0,
           alt_base_uln,
           AST = AST_base,
           ast_uln,
           ALP = ALP_base,
           alp_uln,
           Bili = bili_base,
           bili_uln,
           GGT = GGT_base,
           ggt_uln,
           Platelet = platelet_count_f30080_0_0,
           Albumin = albumin_f30600_0_0,
           AIH,
           PBC,
           AIH.PBC,
           MASLD,
           masld_icd,
           masld_icd_prev,
           masld_icd_inc,
           masld_pdff,
           ALD,
           crypt_sld,
           cir,
           Healthy,
           date_K732,
           date_K738,
           date_K754,
           date_K743,
           date_K746,
           date_I850,
           date_I859,
           date_I982,
           date_I983,
           date_K729,
           date_K703,
           date_K766,
           date_K767,
           enroll_date
           )
```

### Make prevalent and incident cases
AIH cases with an AIH diagnosis (any of 3) before enrollment date are prevalent, rest incident
```{r echo=TRUE, results = 'hide'}
table(df_final$AIH, useNA = "ifany")

# Calculate aih_first_date
df_final$aih_first_date <- pmin(df_final$date_K732, df_final$date_K738, df_final$date_K754, na.rm = TRUE)

# Check the first few values
df_final$aih_first_date[df_final$AIH == TRUE][1:10]

# Now assign NA to aih_first_date where AIH is not TRUE (as some "AIH cases" with concurrent PBC,HBC or HCV were made AIH FALSE initially)
df_final$aih_first_date[df_final$AIH == FALSE] <- NA

# Create the prev_AIH variable
df_final$prev_AIH <- with(df_final, 
                          ifelse(!is.na(aih_first_date) & aih_first_date < enroll_date, 
                                 TRUE, FALSE))

# View the resulting dataframe
table(df_final$prev_AIH, useNA = "ifany")

# Filter the dataframe where prev_AIH is TRUE
df_inspect_prev_AIH <- df_final[df_final$prev_AIH == TRUE, ]

# Visual inspection of the first few rows where prev_AIH == TRUE
print(df_inspect_prev_AIH[, c("eid","AIH", "date_K732", "date_K738", "date_K754", "aih_first_date", "enroll_date", "prev_AIH")])

# Create the inc_AIH variable
df_final$inc_AIH <- with(df_final, 
                         ifelse(!is.na(aih_first_date) & aih_first_date > enroll_date, 
                                TRUE, FALSE))

# View the resulting dataframe
table(df_final$inc_AIH)

# Filter the dataframe where inc_AIH is TRUE
df_inspect_inc_AIH <- df_final[df_final$inc_AIH == TRUE, ]

# Visual inspection of the first few rows where inc_AIH == TRUE
print(df_inspect_inc_AIH[, c("eid","AIH", "date_K732", "date_K738", "date_K754", "aih_first_date", "enroll_date", "inc_AIH")])

df_final %>%
    count(AIH, prev_AIH, inc_AIH)
```

PBC cases with a PBC diagnosis before enrollment date are prevalent, rest incident
```{r echo=TRUE, results = 'hide'}
table(df_final$PBC, useNA = "ifany")
df_final$pbc_date <- df_final$date_K743

# Create the prev_PBC variable
df_final$prev_PBC <- with(df_final, 
                          ifelse(PBC == TRUE & pbc_date < enroll_date, TRUE, FALSE))

# View the resulting dataframe
table(df_final$prev_PBC, useNA = "ifany")

# Filter the dataframe where prev_PBC is TRUE
df_inspect_prev_PBC <- df_final[df_final$prev_PBC == TRUE, ]

# Visual inspection of the first few rows where prev_AIH == TRUE
print(df_inspect_prev_PBC[, c("eid","PBC", "pbc_date", "enroll_date", "prev_PBC")])

# Create the inc_PBC variable
df_final$inc_PBC <- with(df_final, 
                         ifelse(PBC == TRUE & pbc_date > enroll_date, TRUE, FALSE))

# View the resulting dataframe
table(df_final$inc_PBC)

# Filter the dataframe where inc_PBC is TRUE
df_inspect_inc_PBC <- df_final[df_final$inc_PBC == TRUE, ]

# Visual inspection of the first few rows where prev_PBC == TRUE
print(df_inspect_inc_PBC[, c("eid","PBC", "pbc_date", "enroll_date", "inc_PBC")])

df_final %>%
    count(PBC, prev_PBC, inc_PBC)
```

cir cases with a cir diagnosis before enrollment date are prevalent, rest incident
```{r echo=TRUE, results = 'hide'}
# Create the cir_first_date variable to capture the earliest diagnosis date
df_final$cir_first_date <- with(df_final, 
                                pmin(date_K746, date_I850, date_I859, date_I982, date_I983, 
                                     date_K729, date_K703, date_K766, date_K767, na.rm = TRUE))

# Check the output for rows where cir_first_date is not NA
df_inspect_cir <- df_final[!is.na(df_final$cir_first_date), ]
print(head(df_inspect_cir[, c("date_K746", "date_I850", "date_I859", "date_I982", "date_I983", 
                              "date_K729", "date_K703", "date_K766", "date_K767", "cir_first_date")], 10))

# Create the prev_cir variable (TRUE if cir_first_date is before enroll_date)
df_final$prev_cir <- with(df_final, 
                          ifelse(!is.na(cir_first_date) & cir_first_date < enroll_date, TRUE, FALSE))

# Filter the dataframe where prev_cir is TRUE
df_inspect_prev_cir <- df_final[df_final$prev_cir == TRUE, ]

# Visual inspection of the first few rows where prev_cir == TRUE
print(df_inspect_prev_cir[, c("eid","cir_first_date", "enroll_date", "prev_cir")])

# Create the inc_cir variable (TRUE if cir_first_date is after enroll_date)
df_final$inc_cir <- with(df_final, 
                         ifelse(!is.na(cir_first_date) & cir_first_date > enroll_date, TRUE, FALSE))

# Filter the dataframe where inc_cir is TRUE
df_inspect_inc_cir <- df_final[df_final$inc_cir == TRUE, ]

# Visual inspection of the first few rows where inc_cir == TRUE
print(df_inspect_inc_cir[, c("eid","cir_first_date", "enroll_date", "inc_cir")])

df_final %>%
    count(cir, prev_cir, inc_cir)
```

### Ethnicity
```{r echo=TRUE, results='hide'}
# All ethnic groups
df_final$ethnic_all <- df_final$ethnicity
df_final$ethnic_all[df_final$ethnic_all=="Caribbean"] <- 4001
df_final$ethnic_all[df_final$ethnic_all=="Indian"] <- 3001
df_final$ethnic_all[df_final$ethnic_all=="White"] <- 1
df_final$ethnic_all[df_final$ethnic_all=="White and Black Caribbean"] <- 2001
df_final$ethnic_all[df_final$ethnic_all=="British"] <- 1001
df_final$ethnic_all[df_final$ethnic_all=="Pakistani"] <- 3002
df_final$ethnic_all[df_final$ethnic_all=="Mixed"] <- 2
df_final$ethnic_all[df_final$ethnic_all=="African"] <- 4002
df_final$ethnic_all[df_final$ethnic_all=="Irish"] <- 1002
df_final$ethnic_all[df_final$ethnic_all=="White and Black African"] <- 2002
df_final$ethnic_all[df_final$ethnic_all=="Bangladeshi"] <- 3003
df_final$ethnic_all[df_final$ethnic_all=="Asian or Asian British"] <- 3
df_final$ethnic_all[df_final$ethnic_all=="Any other Black background"] <- 4003
df_final$ethnic_all[df_final$ethnic_all=="Any other white background"] <- 1003
df_final$ethnic_all[df_final$ethnic_all=="White and Asian"] <- 2003
df_final$ethnic_all[df_final$ethnic_all=="Any other Asian background"] <- 3004
df_final$ethnic_all[df_final$ethnic_all=="Black or Black British"] <- 4
df_final$ethnic_all[df_final$ethnic_all=="Any other mixed background"] <- 2004
df_final$ethnic_all[df_final$ethnic_all=="Chinese"] <- 5
df_final$ethnic_all[df_final$ethnic_all=="Other ethnic group"] <- 6
df_final$ethnic_all[df_final$ethnic_all=="Do not know"] <- -1
df_final$ethnic_all[df_final$ethnic_all=="Prefer not to answer"] <- -3

#Main ethnic groups (NB this makes some assumptions i.e. British is white, and does not include mixed, Chinese or other Asian)

df_final$ethnic <- df_final$ethnic_all
df_final$ethnic[df_final$ethnic_all==1001] <- 1
df_final$ethnic[df_final$ethnic_all==1002] <- 1
df_final$ethnic[df_final$ethnic_all==1003] <- 1
df_final$ethnic[df_final$ethnic_all==3] <- 2
df_final$ethnic[df_final$ethnic_all==3001] <- 2
df_final$ethnic[df_final$ethnic_all==3002] <- 2
df_final$ethnic[df_final$ethnic_all==3003] <- 2
df_final$ethnic[df_final$ethnic_all==4] <- 3
df_final$ethnic[df_final$ethnic_all==4001] <- 3
df_final$ethnic[df_final$ethnic_all==4002] <- 3
df_final$ethnic[df_final$ethnic_all==4003] <- 3
df_final$ethnic[df_final$ethnic_all==2] <- 4
df_final$ethnic[df_final$ethnic_all==2001] <- 4
df_final$ethnic[df_final$ethnic_all==2002] <- 4
df_final$ethnic[df_final$ethnic_all==2003] <- 4
df_final$ethnic[df_final$ethnic_all==2004] <- 4
df_final$ethnic[df_final$ethnic_all==3004] <- 4
df_final$ethnic[df_final$ethnic_all==5] <- 4
df_final$ethnic[df_final$ethnic_all==6] <- 4
df_final$ethnic[df_final$ethnic_all==-1] <- NA
df_final$ethnic[df_final$ethnic_all==-3] <- NA

count(df_final,ethnic_all) # empty row, should be NA
df_final$ethnic_all # empty row = ""

df_final$ethnic[df_final$ethnic_all==""] <- NA

# 1"White European" 2"South Asian" 3"African Caribbean" 4"Mixed or other"

count(df_final,ethnic)
```

### Clean up dataframe
```{r echo=TRUE, results = 'hide'}
dim(df_final)

df_final <- df_final %>%
  select(-starts_with("date_"), -ethnicity, -ethnic_all)

dim(df_final)
```

### Prevalence estimates
```{r}
# Prevalence estimate (n(diagnosis)/total count)
count(df_final)
table(df_final$AIH, useNA = "ifany")
table(df_final$PBC, useNA = "ifany")
table(df_final$AIH.PBC, useNA = "ifany")
table(df_final$MASLD, useNA = "ifany")
table(df_final$crypt_sld, useNA = "ifany")
table(df_final$ALD, useNA = "ifany")
table(df_final$cir, useNA = "ifany")
table(df_final$Healthy, useNA = "ifany")
```

### Save dataframe
```{r}
summary(df_final)

write.csv(
  df_final,
  here("data/AILD_pheno_UKB_for_proteomics_github.csv"),
  row.names = FALSE
)

```
