---
title: "PBC O-link validation"
author: "JSK"
date: "2026-07-08"
output:
  word_document: default
  html_document: default
  pdf_document: default 
---

```{r echo=T, results='hide', warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(dplyr) 
library(data.table)
library(tableone)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(openxlsx)
```

### Get o-link data
Load Olink proteomics dataset contain normalized protein expression (.NPX) data generated using the antibody-based Olink Explore Proximity Extension Assay (PEA) in blood plasma samples derived from approximately 54,000 UK Biobank participants. These samples (i.e. our dataset) are obtained from participants' baseline visit (between 2006-2010).
https://biobank.ndph.ox.ac.uk/ukb/ukb/docs/PPP_Phase_1_QC_dataset_companion_doc.pdf
```{r echo=T, results='hide', warning=FALSE, message=FALSE}
olink <- fread(here("data-raw/olink_proteins_data_3000.csv"), data.table = FALSE)  # na.strings=c("","NA")
dim(olink) # 53015 individuals, 2923 proteins
olink[1:10,1:10]
olink[53013:53015,1:5]

# Convert column names to uppercase excluding 'eid'
olink <- olink %>%
  rename_with(~ ifelse(. == "eid", ., toupper(.)))
olink[1:10, 1:10]
```

### Get phenotype data
Generated by AILD_pheno_UKB_for_proteomics.Rmd.
Age, Sex, BMI, ethnic, ALT are all values collected at participants initial assessment centre visit. ALT1 is ALT 
level at first repeat assessment visit 2012-13. Diagnoses are TRUE if an individual has got an ICD-10 diagnosis 
of that disease during the period from 1996 to 13.02.2024 (see https://pubmed.ncbi.nlm.nih.gov/37752719/).
```{r echo=T, results='hide', warning=FALSE, message=FALSE}
pheno <-  fread(here("data/AILD_pheno_UKB_for_proteomics.csv"), data.table = FALSE)  # na.strings=c("","NA")
dim(pheno)
pheno[1:10,1:10]
str(pheno)

# Make Sex, alt_base_uln factors
pheno$Sex <- as.factor(pheno$Sex)
table(pheno$Sex, useNA = "ifany")
str(pheno$Sex)
pheno$Sex[pheno$Sex == ""] <- NA
pheno$Sex <- droplevels(pheno$Sex)
table(pheno$Sex, useNA = "ifany")
str(pheno$Sex)

pheno$alp_uln <- as.factor(pheno$alp_uln)
table(pheno$alp_uln, useNA = "ifany")

# Make ethnic a factor with correct labels
pheno$ethnic <- factor(pheno$ethnic, levels = 1:4, labels = c("White European", "South Asian", "African Caribbean", "Mixed or other")) 
table(pheno$ethnic, useNA = "ifany")
dim(pheno)
pheno[1:10,34:38]

# Number of PBC patients regardless of ALP level
table(pheno$PBC, useNA = "ifany") # 403 individuals with PBC

# Make new PBC variable with only those that have increased ALP
pheno$PBC_uln <- with(pheno, ifelse(PBC == TRUE & alp_uln == "elevated", TRUE, FALSE))
pheno$PBC_uln <- as.logical(pheno$PBC_uln)
table(pheno$PBC_uln, useNA = "ifany") #269 índividuals with PBC and elevated ALP at baseline

# Make those PBC_uln NA as FALSE
pheno$PBC_uln[is.na(pheno$PBC_uln)] <- F
table(pheno$PBC_uln, useNA = "ifany")
```

### Merge o-link and pheno dataset
```{r echo=T, results='hide', warning=FALSE, message=FALSE}
# Merge pheno and olink by eid and remove eids not present in both datasets 
olink_df <- merge(pheno,olink,by="eid",ALL=FALSE)
dim(olink_df)
olink_df[1:10,1:50]
```

### Subset dataframe to only include PBC_uln, Healthy and MASLD and make new group variables
```{r echo=T, results='hide', warning=FALSE, message=FALSE}
# Check if any PBC cases in MASLD group, as we don't want to compare with those (some PBC individuals have concurrent MASLD)
olink_df %>%
      count(PBC_uln,PBC,Healthy,MASLD) # 3 PBC with concurrent MASLD

# Subset the dataframe where PBC_uln, Healthy or MASLD is TRUE
olink_subset <- olink_df[
  olink_df$PBC_uln == TRUE | 
  olink_df$Healthy == TRUE | 
  (olink_df$MASLD == TRUE & olink_df$PBC == FALSE),] # As some MASLD cases have concurrent PBC and we don't want to compare our PBC_uln with those

# Now "clean" MASLD group with no PBC individuals
olink_subset %>%
      count(PBC_uln,PBC,Healthy,MASLD) # eids

dim(olink_subset)
dim(olink_df)

# Create a new 'group' variable
olink_subset$group <- with(olink_subset, ifelse(
  PBC_uln == TRUE, "PBC",                   
  ifelse(Healthy == TRUE, "Healthy",        
  ifelse(MASLD == TRUE, "MASLD", NA))))     

# Convert 'group' to a factor
olink_subset$group <- as.factor(olink_subset$group)
table(olink_subset$group, useNA = "ifany")

# Create a new 'group_1' variable with "Con" for Healthy and MASLD, and "PBC" for PBC_uln
olink_subset$group_1 <- with(olink_subset, ifelse(
  PBC_uln == TRUE, "PBC",              
  ifelse(Healthy == TRUE | MASLD == TRUE, "Con", NA)  
))

# Convert 'group_1' to a factor
olink_subset$group_1 <- as.factor(olink_subset$group_1)
table(olink_subset$group_1, useNA = "ifany")
```



### Create Table 1
```{r echo=TRUE, warning=FALSE, message=FALSE}

# Table 1 – INCIDENT PBC CASES ONLY (n = 23)

# Create dataset with incident PBC cases only (controls unchanged)
olink_subset_inc <- olink_subset %>%
  dplyr::filter(group != "PBC" | inc_PBC == TRUE)

# Sanity checks
table(olink_subset_inc$group)
table(olink_subset_inc$inc_PBC[olink_subset_inc$group == "PBC"], useNA = "ifany")

# Variables for Table 1
vari <- c("Age","Sex","BMI","ALT","AST","ALP","Bili","GGT","Platelet",
          "Albumin","hba1c","prev_PBC","inc_PBC","prev_cir","inc_cir","ethnic")

# Create Table 1 (incident cases only)
t1_inc <- tableone::CreateTableOne(
  vars   = vari,
  data   = olink_subset_inc,
  strata = "group"
)

p1_inc <- print(
  t1_inc,
  quote = FALSE,
  noSpaces = TRUE,
  printToggle = FALSE,
  nonnormal = TRUE
)

print(p1_inc)

# Save INCIDENT Table 1
p1_inc <- as.data.frame(p1_inc)
p1_inc$condition <- rownames(p1_inc)
p1_inc <- p1_inc[, c(6,1:5)]

write.xlsx(
  p1_inc,
  here("data/TableOne_PBC_UKB_proteomics_INCIDENT.xlsx"),
  rowNames = TRUE
)

write.csv(
  p1_inc,
  here("data/TableOne_PBC_UKB_proteomics_INCIDENT.csv"),
  row.names = TRUE
)
```


```{r echo=T, results='hide', warning=FALSE, message=FALSE}

# Table 1 prevalent and incident PBC cases (n = 34)

vari <- c("Age","Sex","BMI","ALT","AST","ALP","Bili","GGT","Platelet","Albumin","hba1c","prev_PBC","inc_PBC","prev_cir","inc_cir","ethnic")
t1 <- tableone::CreateTableOne(vars = vari, data = olink_subset, strata = "group")
p1 <- print(t1,quote = FALSE, noSpaces = TRUE, printToggle = FALSE, nonnormal = TRUE)
print(p1)

# write.csv(p1, here("data/TableOne_PBC_UKB_proteomics.csv"))
p1 <- as.data.frame(p1)
p1$condition <- rownames(p1)
p1 <- p1[,c(6,1:5)]
# write.xlsx(p1, here("data/TableOne_PBC_UKB_proteomics.xlsx"), rownames = TRUE)
 
# Check prev_PBC, cir of PBC_uln
olink_subset[olink_subset$PBC_uln == TRUE, c("eid", "prev_PBC", "inc_PBC","prev_cir","inc_cir")] # It is not the one with prev_PBC that has cir

# Chech incident cases of PBC_uln
table(olink_subset$inc_PBC) # 23 individuals with incident PBC and elevated ALP 

# Check prevalence of concurrent MASLD in PBC
table(olink_subset$PBC_uln,olink_subset$MASLD) # 3 but unsure if correct as MASLD prev in UKB is underestimated


# Check diagnosis dates and enrollment date of PBC_uln
olink_subset[olink_subset$PBC_uln == TRUE, c("eid", "enroll_date", "pbc_date","cir_first_date")] # 23 incident and 11 prevalent. 
```

### Check prevalences in olink dataset
```{r}
table(olink_df$PBC, useNA = "ifany") # 52 PBC cases
table(olink_df$PBC_uln, useNA = "ifany") # 34 cases with PBC and elevated ALP
table(olink_df$MASLD, useNA = "ifany") # 6 MASLD cases have concurrent PBC, see below
table(olink_df$PBC, olink_df$MASLD, useNA = "ifany")
table(olink_subset$group, useNA = "ifany")
table(olink_df$Healthy, useNA = "ifany")
```

### Import feature importance proteins from FALL
pro contains list of plasma proteins used for prediction of PBC versus Healthy and MASLD in discovery cohort. 
```{r echo=T, results='hide', warning=FALSE, message=FALSE}
pro <-  fread(here("data-raw/feature_importance_plasma_pbc_vs_healthy_and_masld.csv"), data.table = FALSE) 
dim(pro)
pro
str(pro)

# Check which proteins from pro dataframe are found in olink_subset dataframe 
common_proteins <- pro$Name[pro$Name %in% colnames(olink_subset)]
common_proteins # 12 proteins

# Check which proteins from pro dataframe are not found in olink-subset dataframe
setdiff(pro$Name, colnames(olink_subset)) # 8 proteins
```


### Check missing values in validation proteins

```{r echo=TRUE}
# Count missing values for the proteins included in the validation

missing_common <- sapply(
  olink_subset[, common_proteins],
  function(x) sum(is.na(x))
)

missing_table <- data.frame(
  Protein = common_proteins,
  Missing = missing_common,
  Percent_missing = round(100 * missing_common / nrow(olink_subset), 2)
)

# Show proteins ordered by missingness
missing_table[order(-missing_table$Missing), ]

# Summary statistics
cat("Total missing values:", sum(missing_common), "\n")
cat("Mean percent missing:", round(mean(missing_table$Percent_missing), 2), "%\n")
cat("Maximum percent missing:", round(max(missing_table$Percent_missing), 2), "%\n")
```

### Create boxplots for those importance proteins present in olink dataframe
```{r echo=T, warning=FALSE, message=FALSE}
# Melt the data into long format
olink_long <- olink_subset %>%
  pivot_longer(cols = all_of(common_proteins),  # Select the common proteins
               names_to = "Protein",           # New column for protein names
               values_to = "Expression")       # New column for expression values

# Reorder factor levels so PBC appears first
olink_long$group_1 <- factor(olink_long$group_1, levels = c("PBC", "Con"))

# Create boxplots with statistical comparison
p <- ggplot(olink_long, aes(x = group_1, y = Expression, fill = group_1)) +
  geom_boxplot() +
  labs(title = "Feature Proteins in UKB only PBC cases with elevated ALP (n = 34)", 
       x = "Group", 
       y = "Normalized Protein Expression") +
  theme_minimal() +
  theme(
    legend.position = "none",                     # Hide the default legend
    strip.text = element_text(size = 12, face = "bold")  # Adjust facet label size and make bold
  ) +
  scale_fill_manual(values = c("lightgreen", "lightblue")) +
  stat_compare_means(method = "wilcox.test", label = "p.format") +  
  facet_wrap(~ Protein, scales = "free_y")  # Create facets for each protein

# Print the combined plot
print(p)

# Save boxplot
# ggsave(plot = p, here("data/boxplots_feature_proteins_PBC_UKB.jpeg"), device = "jpeg", width = 9, height = 9, dpi = 300)
```

```{r echo=F, results='hide', warning=FALSE, message=FALSE}
# Just checking that everythings right
tapply(olink_subset$PIGR, olink_subset$group_1,mean,na.rm=TRUE)
tapply(olink_subset$TTR, olink_subset$group_1,mean,na.rm=TRUE)
```

### Create boxplot with only incident PBC cases (with elevated ALP)
```{r echo=T, results='hide', fig.show='hide', warning=FALSE, message=FALSE}
# Subset olink_long to contain only inc_PBC cases
dim(olink_long)
dim(olink_subset)
table(olink_long$group_1,olink_long$prev_PBC)
table(olink_subset$group_1,olink_subset$prev_PBC)

olink_long_no_prev_PBC <- olink_long[olink_long$prev_PBC != TRUE, ]
table(olink_long_no_prev_PBC$group_1,olink_long_no_prev_PBC$prev_PBC)

# Reorder factor levels so PBC appears first
olink_long_no_prev_PBC$group_1 <- factor(olink_long_no_prev_PBC$group_1, levels = c("PBC", "Con"))

# Create boxplots with statistical comparison
p_subset <- ggplot(olink_long_no_prev_PBC, aes(x = group_1, y = Expression, fill = group_1)) +
  geom_boxplot() +
  labs(title = "Feature Proteins in UKB with only incident PBC cases (n = 23)", 
       x = "Group", 
       y = "Normalized Protein Expression") +
  theme_minimal() +
  theme(
    legend.position = "none",                     # Hide the default legend
    strip.text = element_text(size = 12, face = "bold")  # Adjust facet label size and make bold
  ) +
  scale_fill_manual(values = c("lightgreen", "lightblue")) +
  stat_compare_means(method = "wilcox.test", label = "p.format") +  
  facet_wrap(~ Protein, scales = "free_y")  # Create facets for each protein

# Print the combined plot
print(p_subset)

# Save boxplot
# ggsave(plot = p_subset, here("data/boxplots_feature_proteins_incident_PBC_UKB.jpeg"), device = "jpeg", width = 9, height = 9, dpi = 300)
```

### Create boxplot with only non-cirrhotic cases
```{r echo=T, results='hide', fig.show='hide', warning=FALSE, message=FALSE}
# Subset olink_long to contain only non-cirrhotic cases
dim(olink_long)
dim(olink_subset)
table(olink_long$group_1,olink_long$prev_cir)
table(olink_subset$group_1,olink_subset$prev_cir)

olink_long_no_prev_cir <- olink_long[olink_long$prev_cir != TRUE, ]
table(olink_long_no_prev_cir$group_1,olink_long_no_prev_cir$prev_cir)

# Reorder factor levels so PBC appears first
olink_long_no_prev_cir$group_1 <- factor(olink_long_no_prev_cir$group_1, levels = c("PBC", "Con"))

# Create boxplots with statistical comparison
p_subset_no_cir <- ggplot(olink_long_no_prev_cir, aes(x = group_1, y = Expression, fill = group_1)) +
  geom_boxplot() +
  labs(title = "Feature Proteins in UKB with only non-cirrhotic cases (n = 28)", 
       x = "Group", 
       y = "Normalized Protein Expression") +
  theme_minimal() +
  theme(
    legend.position = "none",                     # Hide the default legend
    strip.text = element_text(size = 12, face = "bold")  # Adjust facet label size and make bold
  ) +
  scale_fill_manual(values = c("lightgreen", "lightblue")) +
  stat_compare_means(method = "wilcox.test", label = "p.format") +  
  facet_wrap(~ Protein, scales = "free_y")  # Create facets for each protein

# Print the combined plot
print(p_subset_no_cir)

# Save boxplot
# ggsave(plot = p_subset_no_cir, here("data/boxplots_feature_proteins_no_cir_PBC_UKB.jpeg"), device = "jpeg", width = 9, height = 9, dpi = 300)
```

### Adjust for age and sex
```{r echo=T, results='hide', warning=FALSE, message=FALSE}
# Set 'Con' as the reference level for group_1
olink_subset$group_1 <- relevel(olink_subset$group_1, ref = "Con")

# List to store results for each protein
results_list <- list()

# Loop through each protein in common_proteins
for (protein in common_proteins) {
  # Define the formula for multiple linear regression
  formula <- as.formula(paste(protein, "~ group_1 + Age + Sex"))
  
  # Run the linear regression model
  model <- lm(formula, data = olink_subset)
  
  # Store the summary of the model
  results_list[[protein]] <- summary(model)
}

# Display results for each protein
results_list
```

### Limma analysis of PBC with elevated ALP n=34

```{r echo=TRUE, warning=FALSE, message=FALSE}
library(limma)

# Set control group as reference
olink_subset$group_1 <- relevel(factor(olink_subset$group_1), ref = "Con")
olink_subset$Sex <- factor(olink_subset$Sex)

# Keep samples with complete covariate information
# Missing protein values are kept as NA and are handled protein-wise by limma
limma_df <- olink_subset %>%
  filter(!is.na(group_1),
         !is.na(Age),
         !is.na(Sex),
         !is.na(BMI))

# Create expression matrix: proteins in rows, samples in columns
expr_mat <- limma_df %>%
  select(all_of(common_proteins)) %>%
  as.matrix()

expr_mat <- t(expr_mat)
colnames(expr_mat) <- limma_df$eid

# Design matrix adjusted for age, sex, and BMI
design <- model.matrix(~ group_1 + Age + Sex + BMI, data = limma_df)

# Fit limma model
fit <- lmFit(expr_mat, design)
fit <- eBayes(fit)

# Extract PBC vs controls
limma_ukb_results <- topTable(
  fit,
  coef = "group_1PBC",
  number = Inf,
  adjust.method = "BH",
  sort.by = "P"
)

limma_ukb_results$Protein <- rownames(limma_ukb_results)

# Add available observations per protein and group
n_table <- limma_df %>%
  select(group_1, all_of(common_proteins)) %>%
  pivot_longer(
    cols = all_of(common_proteins),
    names_to = "Protein",
    values_to = "Expression"
  ) %>%
  group_by(Protein, group_1) %>%
  summarise(n_available = sum(!is.na(Expression)), .groups = "drop") %>%
  pivot_wider(
    names_from = group_1,
    values_from = n_available,
    names_prefix = "n_"
  )

# Final results table
limma_ukb_results <- limma_ukb_results %>%
  left_join(n_table, by = "Protein") %>%
  select(Protein, logFC, AveExpr, t, P.Value, adj.P.Val, B, everything())

# Print table in R Markdown output
knitr::kable(limma_ukb_results, digits = 3)

# Save results
write.xlsx(
  limma_ukb_results,
  here("data/limma_UKB_Olink_validation_common_proteins_prevalent_age_sex_BMI_adjusted.xlsx"),
  rowNames = FALSE
)
```
### Limma analysis and boxplot — incident PBC cases only n =23

```{r echo=TRUE, warning=FALSE, message=FALSE}
# Subset to incident PBC cases and controls
olink_subset_incident <- olink_subset %>%
  filter(prev_PBC != TRUE)

# Set control group as reference
olink_subset_incident$group_1 <- relevel(
  factor(olink_subset_incident$group_1),
  ref = "Con"
)

olink_subset_incident$Sex <- factor(olink_subset_incident$Sex)

# Keep samples with complete covariate information
# Missing protein values are kept as NA and handled protein-wise by limma
limma_df_incident <- olink_subset_incident %>%
  filter(!is.na(group_1),
         !is.na(Age),
         !is.na(Sex),
         !is.na(BMI))

# Expression matrix: proteins in rows, samples in columns
expr_mat_incident <- limma_df_incident %>%
  select(all_of(common_proteins)) %>%
  as.matrix()

expr_mat_incident <- t(expr_mat_incident)
colnames(expr_mat_incident) <- limma_df_incident$eid

# Design matrix adjusted for age, sex, and BMI
design_incident <- model.matrix(
  ~ group_1 + Age + Sex + BMI,
  data = limma_df_incident
)

# Fit limma model
fit_incident <- lmFit(expr_mat_incident, design_incident)
fit_incident <- eBayes(fit_incident)

# Extract PBC vs controls
limma_ukb_results_incident <- topTable(
  fit_incident,
  coef = "group_1PBC",
  number = Inf,
  adjust.method = "BH",
  sort.by = "P"
)

limma_ukb_results_incident$Protein <- rownames(limma_ukb_results_incident)

# Add available observations per protein and group
n_table_incident <- limma_df_incident %>%
  select(group_1, all_of(common_proteins)) %>%
  pivot_longer(
    cols = all_of(common_proteins),
    names_to = "Protein",
    values_to = "Expression"
  ) %>%
  group_by(Protein, group_1) %>%
  summarise(n_available = sum(!is.na(Expression)), .groups = "drop") %>%
  pivot_wider(
    names_from = group_1,
    values_from = n_available,
    names_prefix = "n_"
  )

# Final results table
limma_ukb_results_incident <- limma_ukb_results_incident %>%
  left_join(n_table_incident, by = "Protein") %>%
  select(Protein, logFC, AveExpr, t, P.Value, adj.P.Val, B, everything())

# Print table in R Markdown output
knitr::kable(limma_ukb_results_incident, digits = 3)

# Save limma results for incident PBC dataset
write.xlsx(
  limma_ukb_results_incident,
  here("data/limma_UKB_Olink_validation_common_proteins_incident_PBC_age_sex_BMI_adjusted.xlsx"),
  rowNames = FALSE
)

# Create long-format data for the same incident cohort
olink_long_incident <- limma_df_incident %>%
  pivot_longer(
    cols = all_of(common_proteins),
    names_to = "Protein",
    values_to = "Expression"
  )

olink_long_incident$group_1 <- factor(
  olink_long_incident$group_1,
  levels = c("PBC", "Con")
)

# Create labels with adjusted p-values
limma_labels_incident <- limma_ukb_results_incident %>%
  mutate(
    p_label = ifelse(
      adj.P.Val < 0.001,
      "adj. p < 0.001",
      paste0(
        "adj. p = ",
        formatC(adj.P.Val, format = "f", digits = 3)
      )
    )
  ) %>%
  select(Protein, p_label)

# Define y-position for labels in each facet
label_positions_incident <- olink_long_incident %>%
  group_by(Protein) %>%
  summarise(
    y_pos = max(Expression, na.rm = TRUE) +
      0.10 * diff(range(Expression, na.rm = TRUE)),
    .groups = "drop"
  )

limma_labels_incident <- limma_labels_incident %>%
  left_join(label_positions_incident, by = "Protein")

# Create boxplot with limma adjusted p-values
p_subset_limma_adjusted_incident <- ggplot(
  olink_long_incident,
  aes(x = group_1, y = Expression, fill = group_1)
) +
  geom_boxplot() +
  geom_text(
    data = limma_labels_incident,
    aes(x = 1.5, y = y_pos, label = p_label),
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(
    title = "Feature Proteins in UKB with only incident PBC cases",
    x = "Group",
    y = "Normalized Protein Expression"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12, face = "bold")
  ) +
  scale_fill_manual(values = c("lightgreen", "lightblue")) +
  facet_wrap(~ Protein, scales = "free_y")

print(p_subset_limma_adjusted_incident)

# Save boxplot
ggsave(
  plot = p_subset_limma_adjusted_incident,
  filename = here("data/boxplots_feature_proteins_incident_PBC_UKB_limma_adjusted.jpeg"),
  device = "jpeg",
  width = 9,
  height = 9,
  dpi = 300
)


```
### Unadjusted limma analysis of UKB Olink validation data — incident PBC cases only

```{r echo=TRUE, warning=FALSE, message=FALSE}
# Subset to incident PBC cases and controls
olink_subset_incident_unadjusted <- olink_subset %>%
  filter(prev_PBC != TRUE)

# Set control group as reference
olink_subset_incident_unadjusted$group_1 <- relevel(
  factor(olink_subset_incident_unadjusted$group_1),
  ref = "Con"
)

# Keep samples with group information
# Missing protein values are kept as NA and handled protein-wise by limma
limma_df_incident_unadjusted <- olink_subset_incident_unadjusted %>%
  filter(!is.na(group_1))

# Expression matrix: proteins in rows, samples in columns
expr_mat_incident_unadjusted <- limma_df_incident_unadjusted %>%
  select(all_of(common_proteins)) %>%
  as.matrix()

expr_mat_incident_unadjusted <- t(expr_mat_incident_unadjusted)
colnames(expr_mat_incident_unadjusted) <- limma_df_incident_unadjusted$eid

# Unadjusted design matrix
design_incident_unadjusted <- model.matrix(
  ~ group_1,
  data = limma_df_incident_unadjusted
)

# Fit limma model
fit_incident_unadjusted <- lmFit(
  expr_mat_incident_unadjusted,
  design_incident_unadjusted
)

fit_incident_unadjusted <- eBayes(fit_incident_unadjusted)

# Extract PBC vs controls
limma_ukb_results_incident_unadjusted <- topTable(
  fit_incident_unadjusted,
  coef = "group_1PBC",
  number = Inf,
  adjust.method = "BH",
  sort.by = "P"
)

limma_ukb_results_incident_unadjusted$Protein <- rownames(
  limma_ukb_results_incident_unadjusted
)

# Add available observations per protein and group
n_table_incident_unadjusted <- limma_df_incident_unadjusted %>%
  select(group_1, all_of(common_proteins)) %>%
  pivot_longer(
    cols = all_of(common_proteins),
    names_to = "Protein",
    values_to = "Expression"
  ) %>%
  group_by(Protein, group_1) %>%
  summarise(n_available = sum(!is.na(Expression)), .groups = "drop") %>%
  pivot_wider(
    names_from = group_1,
    values_from = n_available,
    names_prefix = "n_"
  )

# Final results table
limma_ukb_results_incident_unadjusted <- limma_ukb_results_incident_unadjusted %>%
  left_join(n_table_incident_unadjusted, by = "Protein") %>%
  select(Protein, logFC, AveExpr, t, P.Value, adj.P.Val, B, everything())

# Print table
knitr::kable(limma_ukb_results_incident_unadjusted, digits = 3)

# Save results
write.xlsx(
  limma_ukb_results_incident_unadjusted,
  here("data/limma_UKB_Olink_validation_common_proteins_incident_PBC_unadjusted.xlsx"),
  rowNames = FALSE
)
```
